---
title: 'Waller SISMID 2023 Reading and Mapping Shapefiles:  Alcohol, Drugs, and Crime in Houston'
author: "Lance A. Waller"
date: "7/18/2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

***

* **What we have**
An ArcGIS shapefile for census tracts in Houston.
+ A "shapefile" is actually several separate files with the same file name and different extentions.
+ The .dbf file is a dbase database file containing the attribute table
+ The .shp file contains information on the outlines of the tracks.
+ The .shx, .sbn, and .sbx files have the connective tissue to link it all together (sometims you can have a .prj paper for the map projection).
The following attributes are of interest to us:  Population (census), violent crimes (police), number of illegal drug arrests (police), total alcohol sales (state alcoholic beverage license data)

***

* **What we want**
+ *Choropleth* maps of the violent crime rate and standardized log drug arrests and standardized log alcohol sales.
+ Will need to calculate these variables, choose intervals and colors, and map.

***

* **How we get it**
+ We will use the 'rgdal' package and the function 'readOGR' function.
+ We will use the 'RColorBrewer' and 'classInt' libraries to choose colors and intervals.
+ For this example, we will use the basic 'plot' function to make maps.

***

First, load the libraries. The working directory should automatically be set to the correct one if you opened the R project (`2023-SISMID-Spatial-Epi.Rproj`) through RStudio.  

```{r, warning=FALSE}
##Load libraries

library(maptools)     # loads sp library too
library(RColorBrewer) # creates nice color schemes
library(classInt)     # finds class intervals for continuous variables
library(spgwr)        # Adds the geographically weighted regression functions
library(rgdal)        # Provides tool for reading in shapefiles
library(here)         # For constructing filepaths relative to root directory
```
***

Now to read in the shapefile.  The following five files all constitute what is a "shapefile." If you downloaded the repository from GitHub correctly, they should all be in your `data` folder. 

+ HoustonENAR2012final.shp
+ HoustonENAR2012final.dbf
+ HoustonENAR2012final.shx
+ HoustonENAR2012final.sbx
+ HoustonENAR2012final.sbn

```{r}
##Read in shapefile - Houston Census Tracts
houston = readOGR(dsn = here("data"),layer = "HoustonENAR2012final")
```
***

## Plotting the map
+ To get the outlines simply plot our 'houston' data object.

***
```{r}
plot(houston)
```
***

Next, we want to make choropleth maps (shading in each tract based on it attribute value).

To do this, we need to decide how many intervals (colors) we want.  I like odd numbers so there is a 'middle' color, and I usually start with quintiles.

'classInt' will assign each tract to the appropriate quintile for a particular attribute.

'RColorBrewer' will assign a color scheme for the quantiles...lots of fascinating work on color choices by Cynthia Brewer, well worth reading and checking the ColorBrewer webpage (https://colorbrewer2.org/)

```{r}
## Making choropleth maps
# Plot the outlines (we'll add color below)
plot(houston)

# Define the variable (attribute) to shade tracts by
pop2000 <- houston@data$POP2000

# Define the number of classes
nclr <- 5  # quintiles

# Use RColorBrewer to choose the colors
plotclr <- brewer.pal(nclr,"BuPu")
class <- classIntervals(pop2000, nclr, style="quantile")
colcode <- findColours(class, plotclr)

#Fill in the tracts with the colors
plot(houston, col=colcode, add=T)
#Add a title
title(main="Population 2000",
      sub="Quantile (Equal-Frequency) Class Intervals")

#Add a legend  (Coordinates are in longitude, latitude).
legend(-95.7, 29.65, legend=names(attr(colcode, "table")),
       fill=attr(colcode, "palette"), cex=0.6, bty="n")
```

***

OK, great, we can map the data.  Let's map the main variables for our analysis.

The data table has a lot of census data and various transformations of the violent crime, alcohol sales, and drug arrest data.  The next section pulls the values we want.
***
```{r}
# Outcome:  Number of violent crimes by tract
violence = houston@data$violence_2

# Divide by the 2000 population to get the rate
violence.rate = violence/houston@data$tot_pop

# Covariate 1 (log standardized total alcohol sales)
Z.log.total = houston@data$Zl_total

# Covariate 2 (log standardized illegal drug arrests)
Z.log.drug = houston@data$Zl_drug
```
***
Now to map the outcome.
***

```{r}

# Plot Outcome first
plot(houston)
# Define the number of classes
nclr <- 5  # quintiles
# Use RColorBrewer to choose the colors
plotclr <- brewer.pal(nclr,"BuPu")
class <- classIntervals(violence.rate, nclr, style="quantile")
colcode <- findColours(class, plotclr)
#Fill in the tracts with the colors
plot(houston, col=colcode, add=T)
#Add a title
title(main="Violence Rate",
      sub="Quantile (Equal-Frequency) Class Intervals")
#Add a legend  (Coordinates are in longitude, latitude).
legend(-95.7, 29.65, legend=names(attr(colcode, "table")),
       fill=attr(colcode, "palette"), cex=0.6, bty="n")
```
***
Next, map standardized log total alcohol sales.
***

```{r}
# Next plot stdized log total alcohol sales
plot(houston)
# Define the number of classes
nclr <- 5  # quintiles
# Use RColorBrewer to choose the colors
plotclr <- brewer.pal(nclr,"BuPu")
class <- classIntervals(Z.log.total, nclr, style="quantile")
colcode <- findColours(class, plotclr)
#Fill in the tracts with the colors
plot(houston, col=colcode, add=T)
#Add a title
title(main="Std log total sales",
      sub="Quantile (Equal-Frequency) Class Intervals")
#Add a legend  (Coordinates are in longitude, latitude).
legend(-95.7, 29.65, legend=names(attr(colcode, "table")),
       fill=attr(colcode, "palette"), cex=0.6, bty="n")
```
***
Finally, plot standardized log illegal arrests.
***

```{r}
# Next plot stdized log illegal arrests
plot(houston)
# Define the number of classes
nclr <- 5  # quintiles
# Use RColorBrewer to choose the colors
plotclr <- brewer.pal(nclr,"BuPu")
class <- classIntervals(Z.log.drug, nclr, style="quantile")
colcode <- findColours(class, plotclr)
#Fill in the tracts with the colors
plot(houston, col=colcode, add=T)
#Add a title
title(main="Std log drug arrests",
      sub="Quantile (Equal-Frequency) Class Intervals")
#Add a legend  (Coordinates are in longitude, latitude).
legend(-95.7, 29.65, legend=names(attr(colcode, "table")),
       fill=attr(colcode, "palette"), cex=0.6, bty="n")

```
***

These three figures will match the maps in Figure 1 of:  

Waller LA, Zhu L, Gotway CA, Gorman DM, and
Gruenewald PJ (2007) "Quantifying geographic variations in
associations between alcohol distribution and violence: A
comparison of geographically weighted regression and spatially
varying coefficient models".  *Stochastic
Environmental Research and Risk Assessment*. **21**, 573-588.
***
